# Datasets {#datasets}

This guide uses the **QIIME2 Moving Pictures** demo dataset.

The data are downloaded as QIIME2 artifacts and converted into a single
`phyloseq` object using reproducible scripts:

-   `scripts/00-download-qiime2-demo-data.R`
-   `scripts/01-create-phyloseq-object.R`

Those scripts:

-   download the feature table, taxonomy, and metadata
-   align sample identifiers
-   construct a validated `phyloseq` object
-   save a canonical `.rds` file for reuse across chapters

From this point onward, we work with one structured object.

## Load the canonical object

``` {r}
ps <- readRDS("data/moving-pictures-ps.rds")
ps
```

::: cdi-message
A microbiome dataset is not a single table.

It is a coordinated structure containing: - an abundance matrix -
sample-level metadata - taxonomic annotation

All components must align exactly.
:::

## Inspect the OTU Table

``` {r}
otu <- as(otu_table(ps), "matrix")

if (!taxa_are_rows(ps)) {
  otu <- t(otu)
}

otu[1:6, 1:6]
```

::: cdi-message
This is a small slice of the count matrix.

-   Rows represent taxa (ASVs)
-   Columns represent samples
-   Values are raw read counts

Most entries are zero.

Sparsity is not a technical artifact --- it is a biological reality.
:::

## Inspect Sample Metadata

``` {r}
head(sample_data(ps))
```

::: cdi-message
Metadata provides the biological and experimental context.

Without metadata, abundance values have no meaning. Visualization is
always performed relative to metadata.
:::

## Inspect Taxonomy

``` {r}
head(tax_table(ps))
```

::: cdi-message
Taxonomy links sequence variants to biological hierarchy.

Aggregation (for example genus-level barplots) depends entirely on this
mapping. If taxonomy is inconsistent, interpretation becomes unstable.
:::

## Structural Integrity

``` {r}
all(sample_names(ps) == rownames(sample_data(ps)))
```

::: cdi-message
A valid `phyloseq` object guarantees alignment across components.

Many downstream errors originate from mismatched sample identifiers.
:::

:::next
Next → [Data Structure and Prevalence](#data-structure-prevalence)
:::
