# Data structure and prevalence

Microbiome count data is structurally sparse and uneven.

This chapter focuses on:

1.  Matrix sparsity\
2.  Library size variation\
3.  Taxa prevalence

## Load data

``` {r}
library(phyloseq)
ps <- readRDS("data/moving-pictures-ps.rds")
```

## Sparsity

``` {r}
otu <- as(otu_table(ps), "matrix")
if (!taxa_are_rows(ps)) otu <- t(otu)

zero_prop <- mean(otu == 0)
zero_prop
```

::: cdi-message
A high proportion of zeros is normal in microbiome data.

Filtering decisions are not cosmetic. They determine which patterns
become visible.
:::

## Library Size Variation

``` {r}
lib_size <- sample_sums(ps)
summary(lib_size)
```

``` {r}
library(ggplot2)

df_lib <- data.frame(
  sample_id = names(lib_size),
  library_size = as.numeric(lib_size)
)

ggplot(df_lib, aes(x = library_size)) +
  geom_histogram(bins = 20) +
  labs(x = "Library size", y = "Number of samples")
```

::: cdi-message
When sequencing depth varies, raw counts are not comparable.

Normalization is required for fair comparison across samples.
:::

## Detection and Prevalence

``` {r}
detection <- 1

taxa_prev <- rowSums(otu >= detection)
taxa_prev_df <- data.frame(
  taxon = rownames(otu),
  prevalence_n = taxa_prev,
  prevalence_frac = taxa_prev / ncol(otu)
)

summary(taxa_prev_df$prevalence_n)
```

``` {r}
ggplot(taxa_prev_df, aes(x = prevalence_n)) +
  geom_histogram(bins = 20) +
  labs(x = "Number of samples detected", y = "Number of taxa")
```

::: cdi-message
Most taxa are observed in only a small fraction of samples.

Prevalence filtering improves clarity, but it also changes the feature
space. Filtering rules must be stated explicitly.
:::

## Simple Prevalence Filter

``` {r}
min_prev_frac <- 0.10
keep_taxa <- taxa_prev_df$taxon[taxa_prev_df$prevalence_frac >= min_prev_frac]

ps_prev <- prune_taxa(keep_taxa, ps)

c(original_taxa = ntaxa(ps), after_filter = ntaxa(ps_prev))
```

::: cdi-message
Prevalence filtering reduces dimensionality and improves readability.

It can also change ordination and diversity results. Interpretation must
acknowledge these choices.
:::

## Relative Abundance Preview

``` {r}
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
range(sample_sums(ps_rel))
```

::: cdi-message
After transformation, each sample sums to 1.

This enables clearer visualization, but does not remove compositional
constraints.
:::

:::next
Next → [Composition Visualization](#composition-visualization)
:::
