# Composition Visualization

Composition plots are often the first figures shown in microbiome
studies.

Bar plots at the phylum or genus level are common. They appear
intuitive. They are also easy to misinterpret.

::: cdi-message
Relative abundance plots do not show absolute change.

They show proportions constrained to sum to one.
:::

This chapter focuses on:

-   transforming counts to relative abundance
-   aggregating taxa to a chosen rank
-   visualizing composition across metadata groups
-   understanding compositional limitations

## Load Data

``` {r}
library(phyloseq)
library(ggplot2)

ps <- readRDS("data/moving-pictures-ps.rds")
```

## Transform to Relative Abundance

Raw counts are not comparable across samples with different sequencing
depths.

``` {r}
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))

range(sample_sums(ps_rel))
```

::: cdi-message
Each sample now sums to 1.

This enables proportional comparison across samples. However, relative
abundance introduces compositional constraints. An increase in one taxon
must decrease another.
:::

## Aggregate to Genus Level

``` {r}
ps_genus <- tax_glom(ps_rel, taxrank = "Genus")

ntaxa(ps_genus)
```

Remove taxa without genus assignment:

``` {r}
ps_genus <- subset_taxa(ps_genus, !is.na(Genus))
ntaxa(ps_genus)
```

::: cdi-message
Aggregation simplifies visualization, but it also hides within-genus
variation.

Resolution is always a trade-off.
:::

## Identify Top Genera

To improve readability, select the globally most abundant genera.

``` {r}
genus_abundance <- taxa_sums(ps_genus)

top_genera <- names(sort(genus_abundance, decreasing = TRUE))[1:10]

ps_top <- prune_taxa(top_genera, ps_genus)

ntaxa(ps_top)
```

## Prepare Data for Plotting

``` {r}
df_comp <- psmelt(ps_top)
```

## Composition by Body Site

``` {r}
ggplot(df_comp, aes(x = Sample, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ `body-site`, scales = "free_x") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(y = "Relative Abundance")
```

::: cdi-message
Certain genera dominate specific body sites.

But dominance in a relative abundance plot does not imply absolute
biological increase.
:::

## Group-Level View

``` {r}
ggplot(df_comp, aes(x = `body-site`, y = Abundance, fill = Genus)) +
  geom_boxplot() +
  labs(y = "Relative Abundance")
```

::: cdi-message
Boxplots help reveal distribution spread and variability.

Composition plots are descriptive, not inferential.
:::

## Limitations of Composition Plots

Composition plots:

-   are constrained by constant-sum scaling
-   cannot reveal absolute abundance changes
-   are sensitive to filtering and aggregation level

::: cdi-message
Composition plots answer: "What proportion of the community does each
taxon occupy?"

They do not answer: "Did the taxon increase in absolute abundance?"
:::

::: next
Next → [Diversity Analysis](#diversity-analysis)
:::
